Do LSTMs Understand the Licensing of Negative Polarity Items?

### 2021-12-21 paper.tex
Updates from Overleaf

The development of Recurrent Neural Networks (RNN) has been a recent hallmark of advancements in Natural Language Processing (NLP). The inherent ability for RNNs to learn and internalize a number of linguistic capabilities is evidenced by a large corpus of research in the utilization of RNNs for NLP tasks such as machine translation, language modeling, and syntactic parsing \cite{wilcox-etal-2018-rnn}. However, it is still largely unknown how such networks are able to deduce linguistic information, in that there is comparatively less literature present regarding if networks' success is due to opportunistic surface-pattern-based heuristics or if networks are able to genuinely extract and interpret structural properties, syntactic or semantic, of language from their training data \cite{lakretz2019emergence}.

One subset of RNNs, namely Long Short-Term Memory RNNs (LSTMs), has grown in utility in said extraction of structural properties of language. Wilcox et al. \shortcite{wilcox-etal-2018-rnn} determined utilized LSTM models to explore the syntactic relationship of the filler-gap dependency, the relationship between a filler word (in English, a wh- complementizer like "what" or "when") and a gap word (in English, an empty syntactic position uniquely licensed by a filler word). Lakretz et al. \shortcite{lakretz2019emergence} similarly utilized LSTM models for exploring numerical agreement between verbs and their arguments.

There is, though, little research investigating the feature of licensing of negative polarity items in language. Negative polarity items are a class of words that are only permitted to appear in certain “negative” contexts. For example, in English, negative polarity items include words like "ever," "any," and "even."
\begin{enumerate}
    \item He never gave up.
    \item *He \textit{ever} gave up.
    \item Nobody ever gave up.
\end{enumerate}
Here, the word "ever" is not permitted to occur in a sentence except when licensed by a word such as "nobody" that provides the surrounding "negative" context. Notably, the relationship between the licensor and the negative polarity item is based on the hierarchical structure of a sentence, wherein the licensor must c-command the negative polarity item (i.e. the negative polarity item must be a sister of the licensor or a direct descendant of a sister node of the licensor).

>>>>

The development of Recurrent Neural Networks (RNN) has been a recent hallmark of advancements in Natural Language Processing (NLP). The inherent ability for RNNs to learn and internalize a number of linguistic capabilities is evidenced by a large corpus of research in the utilization of RNNs for NLP tasks such as machine translation, language modeling, and syntactic parsing \cite{wilcox-etal-2018-rnn}.

However, it is still largely unknown how such networks are able to deduce linguistic information, in that there is comparatively less literature present regarding if networks' success is due to opportunistic surface-pattern-based heuristics or if networks are able to genuinely extract and interpret structural properties, syntactic or semantic, of language from their training data \cite{lakretz2019emergence}.

One subset of RNNs, namely Long Short-Term Memory RNNs (LSTMs), has grown in utility in said extraction of structural properties of language. Wilcox et al. \shortcite{wilcox-etal-2018-rnn} determined utilized LSTM models to explore the syntactic relationship of the filler-gap dependency, the relationship between a filler word (in English, a wh- complementizer like \textit{what} or \textit{when}) and a gap word (in English, an empty syntactic position uniquely licensed by a filler word). Lakretz et al. \shortcite{lakretz2019emergence} similarly utilized LSTM models for exploring numerical agreement between verbs and their arguments.

There is, though, little research investigating the feature of licensing of negative polarity items in language. Negative polarity items are a class of words that are only permitted to appear in certain "negative" contexts. For example, in English, negative polarity items include words like \textit{ever}, \textit{any}, and \textit{even}.
\begin{enumerate}
    \item He never gave up.
    \item *He \textit{ever} gave up.
    \item Nobody ever gave up.
\end{enumerate}
Here, the word \textit{ever} is not permitted to occur in a sentence except when licensed by a word such as \textit{nobody} that provides the surrounding "negative" context. Notably, the relationship between the licensor and the negative polarity item is based on the hierarchical structure of a sentence, wherein the licensor must c-command the negative polarity item (i.e. the negative polarity item must be a sister of the licensor or a direct descendant of a sister node of the licensor).
